Evaluating MHD parameters of relativistic shock waves with particle-in-cell modeling

### 2020-08-22 papers/Physica2020/Romansky_Physica2020.tex
paper

\documentclass[a4paper]{jpconf}
\bibliographystyle{iopart-num}
\usepackage{amsmath}
\usepackage{citesort}
\usepackage{subfigure}
\usepackage{graphicx}
\graphicspath{{fig/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi
\usepackage[export]{adjustbox}

@@

\begin{document}
\title{Evaluating MHD parameters of relativistic shock waves with Particle-in-Cell modelling}

\author{V I Romansky$^{1}$, A M Bykov$^{1,2}$ and S M Osipov$^1$}

@@

\begin{abstract}
Reklativistic shock waves are widely discussed phenomenae of modern astrophysics. Studying of their properties is necessary for consistent models of supernova remnants, jets and other astrophysical objects. But magneto-hydrodinamical description of relativistic shock waves is not full, because it does not take into account electrons and a lot of relativistic effects such as displacement current density and changing of adiabatic index That's why we need more accurate methods to check and fix MHD models.
\end{abstract}
\section{Introduction}
Relativistic shock jump (Rankine-Hugoniot) were given by Blandford and McKee \cite{Blandford76} and generalized to MHD by Emmering and Chevalier\cite{Emmering87}. Shock conditions are:
\begin{equation}\label{hugoniot}
[\![\rho u^r]\!] = [\![ (w + \frac{B^2}{4 \pi})u^r u^t]\!] = [\![(w + \frac{B^2}{4 \pi}){u^r}^2 + P + \frac{B^2}{8 \pi}]\!] = [\![\frac{B}{\rho}]\!] = 0
\end{equation}
where $u = (u^t,u^r)$ is velocity four-vector measured in the shock frame, $\rho$ is proper mass density, $B$ is magnetic field and $P$ is pressure in the fluid frame and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is enthalpy, $\hat{\gamma}$ is adiabatic index and double brackets indicate the mismatches of quantities to both sides of the shock.

In more simple case of strictly perpendicular shock, shock velocity and wownstream temperature can be expressed\cite{Amato2006} as
\begin{equation}\label{vshock}
v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace
\end{equation}
\begin{equation}\label{temperature}
T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh})
\end{equation}
Here $\sigma$ is magnetisation and $\Gamma$ is upstream lorentz factor measured in downstream frame.

One weak point of this formalism is adiabatic index $\gamma$. For monoatomic gas it is well-defined in classical limit $\gamma = \frac{5}{3}$ and in ultra-relativistic linit $\gamma = \frac{4}{3}$. In the gap between them it is a non-trivial question which value of $\gamma$ to choose. Another problem is that one-fluid model doesn't take into account the difference between electrons' and protons' temperature. And two-fluid model also use some phenomenological methods to evaluate electrons' temperautre.

In this work we used Particle-in-Cell method for modelling relativistic shock waves to study accuracy of MHD equations and to determine adiabatic index.

\section{Numerical setup}
We simulated a structure of a perpendicular relativistic shock and varied magnetisation, lorentz-factor and fraction of turbulent field. 
The simulation is two dimensional with three dimensional velocities and fields. For the numerical simulation we used Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.

Shock wave is initialized by an electron-proton plasma which is flowing into the simulation region through the right boundary and then reflecting from the super-conducting wall at the left boundary. We use the following parameters for the setups: the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.04$ (in the turbulent case $B^2$ is the mean square field). The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal $10^{-4}$ and the proton mass is reduced to $m_p = 100 m_e$. Size of the simulation region along x axis is $L_x = 30000\frac{c}{\omega_p}$ and in transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p$ is the plasma frequency $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$. These sizes correspond to $150000$ and $2000$ grid points respectively. Also, $2000$ grid points correspond to approximately $10$ gyroradii of upstream protons.
In simulation with turbulent fields, we initialized it as summ of harmonic modes with Kolmogorov's spectum, and it is described in our previous paper in detail\cite{Romansky2019}. The energy density of turbulent field compared to the total magnetic energy density is $\eta$ and varying from 0 percent to 90 percent.

\section{Results}
Using the density profiles of shock wave, we derived magnetohydrodynamic parameters of plasma, using formulae \ref{vshock} and \ref{temperature}. Density, averaged in transversal directions  has clear jump as shown on figures \ref{density_noturb} and \ref{density_turb}, indicating the coordinate of shock front.

\begin{figure}[h!]
	\centering
	\begin{minipage}{0.5\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
		\caption{Averaged by y dimension density profile of shock wave.}
		\label{density_noturb}
	\end{minipage}\hfill
	\begin{minipage}{0.5\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
		\caption{Averaged by y dimension density profile of shock wave in turbulent medium.}
		\label{density_turb}
	\end{minipage}
\end{figure}

Evalation of shock front coordinates at different time moments shows that shock front velocity is constant with high accuracy in the case without turbulent field and has small oscillations without turbulence, see figure\ref{shockx}.
\begin{table}
	\label{setups}
	\begin{center}
		\begin{tabular}{|c| c| c| c| c| c| c| c| c| c| c| c|}
		\hline
			Setup & Nx & Ny & Nz & $\sigma$ & $\Gamma$ & $\eta$ & $\beta$ & $\hat{\gamma}$ & $T  10^{12} K$ & $T_p  10^{12} K$ & $T_e  10^{12} K$\\
			reg1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 0 & 0.394 & 1.349 & 6.38 & 1 & 1\\
			reg2 & 150000 & 2000 & 1 & 0.004 & 1.5 & 0 & 0.355 & 1.35 & 5.79 & 1 & 1\\
			reg3 & 150000 & 2000 & 1 & 0.0004 & 1.5 & 0 & 0.339 & 1.338 & 5.53 &1 & 1\\
			turb1 & 150000 & 2000 & 1 & 0.04 & 1.5 & 90 & 0.309 & 1.247 & 5.0 & 1 & 1\\
			turb2 & 150000 & 2000 & 1 & 0.04 & 1.5 & 30 & 0.356 & 1.305 & 5.77 & 1 & 1\\
			turb3 & 150000 & 2000 & 1 & 0.004 & 1.5 & 90 & 0.309 & 1.303 & 5.03 & 1 & 1\\
			turb4 & 150000 & 2000 & 1 & 0.004 & 1.5 & 30 & 0.348 & 1.343 & 5.67 & 1 & 1\\
			gam1 & 150000 & 2000 & 1 & 0.04 & 2 & 0 & 0.447 & 1.411 & 9.67 & 1 & 1\\
			gam2 & 150000 & 2000 & 1 & 0.04 & 5 & 0 & 0.505 & 1.475 & 27.3 & 1 & 1\\
			gam3 & 150000 & 2000 & 1 & 0.04 & 10 & 0 & 0.528 & 1.499 & 57.1 & 1 & 1\\
			reg3d & 80000 & 40 & 40 & 0.04 & 1.5 & 0 & 0.447 & 1.43 & 1 & 1 & 1\\

		\hline
		\end{tabular}
	\end{center}
	\caption{Parameters of different setups}
\end{table}

\begin{figure}[h!]
	\centering
	\includegraphics[width=0.6\textwidth]{fig/shock_x.png} 
	\caption{Coordinate of shock front for different setups.}
	\label{shock_x}
\end{figure}

\section{Conclusions}
In this paper we presented the results of particle-in-cell simulation of electron acceleration in trans-relativistic shock propagating in a clumped  wind of a massive star. Simulations show that the presence  of background  magnetic turbulence in the clumped wind has strong impact on particle acceleration and the acceleration is efficient for high enough level of the magnetic turbulence. The electron acceleration is highly suppressed in the case of  trans-relativistic shock propagating quasi-perpendicular to a regular magnetic field in the wind without the background magnetic turbulence. With the PIC simulations we demonstrated here that the presence of magnetic fluctuations in the background wind results in efficient electron acceleration by the quasi-perpendicular trans-relativistic shocks. We find that the scales and amplitudes of the magnetic turbulent field are important for this process. Further PIC simulations with a wider dynamical range will allow to study the effect of turbulent wind on particle acceleration in detail.

\ack
The authors acknowledge a support from RSF grant 16-12-10225.

%\newpage

\section*{References}

\bibliographystyle{iopart-num}
\bibliography{bibliogr}
\end{document}

>>>>

\documentclass[a4paper]{jpconf}
% \bibliographystyle{iopart-num}
\usepackage{amsmath}
\usepackage{citesort}
\usepackage{subfigure}
\usepackage{graphicx}
\graphicspath{{fig/}}
\usepackage{ifpdf}
\ifpdf\usepackage{epstopdf}\fi
\usepackage[export]{adjustbox}

@@

\usepackage{xspace}
\usepackage[dvipsnames]{xcolor}
\usepackage[normalem]{ulem}
% Suggest to remove
\newcommand{\asrm}[1]{{\color{OrangeRed}\sout{#1}}}
% Suggest to insert
\newcommand{\as}[1]{\color{RoyalBlue}#1\xspace\color{black}}
% Suggest to replace
\newcommand{\asrp}[2]{\asrm{#1} \as{#2}}
% Comment
\newcommand{\ascm}[1]{{\color{ForestGreen}#1}}

\begin{document}
	\title{Evaluating MHD parameters of relativistic shock waves with particle-in-cell modeling}

	\author{V I Romansky$^{1}$, A M Bykov$^{1}$ and S M Osipov$^{1,2}$}

@@

	\begin{abstract}
		Relativistic plasma outflows are observed in gamma-ray burst 
		sources, jets of active galactic nuclei, pulsar wind nebulae and
		supernovae explosions. Magnetohydrodynamical (MHD) shock waves
		inevitably result from interactions of such relativistic outflows with
		the ambient interstellar matter. The widely used single-fluid MHD
		description of relativistic shock waves is a useful tool to study the
		global structure of such objects. However, to interpret the observed
		emission spectra of space objects with relativistic shocks, a kinetic
		description of electrons, positrons, and ions is needed. The kinetic
		structure of relativistic shocks can be modelled within the
		particle-in-cell approach, which allows one to properly account for
		the electron scale processes and for the density of the displacement
		current. Here we discuss the effect of the micro-scale plasma processes on such
		macroscopic parameters of relativistic shocks as the adiabatic index
		of the relativistic fluid in the shock vicinity, with a relatively
		long equilibration time. The adiabatic index is an important parameter of the
		single-fluid MHD models commonly used for shock modeling at much
		longer hydrodynamical scales.
	\end{abstract}

	\section{Introduction}
	Relativistic jets and winds are the generic outflows of the relativistic supernovae \cite{2010Natur.463..513S,2007ApJ...667..351W}, the accreting black holes, both suprmassive in the active galactic nucleis  \cite{1984RvMP...56..255B} as well as the stellar masses in the gamma-ray burst sources, microquasars \cite{2019MmSAI..90...57M,1999PhR...314..575P,2014LNP...876.....R} and young pulsars \cite{2017SSRv..207....1B}. Relativistic shocks can be formed either inside the flow of low or moderate magnetization (the internal shocks) or can be driven by the outflow colliding with the ambient medium (the external shocks). In most cases the relatistivistic shocks in astrophysical objects are collisionless and are associated with strongly non-equilibrium particle distributions
	\cite{2012SSRv..173..309B,2015SSRv..191..519S,2017SSRv..207..319P}, and they can accelerate particles to ultra-high energies \cite{2009JCAP...11..009L}. On the one hand, the presence of non-maxwellian components with very long equilibration time requires a kinetic description of the shocked flows. On the other hand, available computer capabilities do not alow one to provide a fully kinetic modelling of the macroscopic astrophysical flows and, therefore, the fluid MHD description based on such macroscopic parameters as the effective adiabatic index of the flow is widely used. Here we discuss an approach allowing to determine the effective adiabatic indexes from the microscopical particle-in-cell modeling of the relativistic shocks.

	Relativistic fluid shock dynamics was discussed by Lichnerowicz \cite{1967rhm..book.....L}, Blandford and McKee \cite{Blandford76} and by Emmering and Chevalier\cite{Emmering87}. The relativistic shock jump conditions in the case of a perpendicular shock which is typical for relativisic shocks can be written as:
	\begin{equation}\label{hugoniot}
	[\![\rho u^r]\!] = [\![ (w + \frac{B^2}{4 \pi})u^r u^t]\!] = [\![(w + \frac{B^2}{4 \pi}){u^r}^2 + P + \frac{B^2}{8 \pi}]\!] = [\![\frac{B}{\rho}]\!] = 0
	\end{equation}
	where $u = (u^t,u^r)$ is the velocity four-vector measured in the shock frame, $\rho$ is the proper mass density, $B$ is the magnetic field, $P$ is the pressure in the fluid frame, and $w = \rho c^2 + \frac{\hat{\gamma}}{\hat{\gamma} - 1} P$ is the enthalpy, while $\hat{\gamma}$ is the adiabatic index and the double brackets indicate the jump of the quantities between two sides of the shock.

	In the ultra-relativistic limit, the shock velocity $v_{sh}$ and the downstream temperature $T$ can be expressed as, respectively, \cite{Amato2006,2011A&ARv..19...42B}:
	\begin{equation}\label{vshock}
	v_{sh} = \frac{c}{2}\frac{1}{1 + \sigma}\lbrace (\hat{\gamma}(1 + \frac{\sigma}{2})-1) + {\lbrack (\hat{\gamma}(1 + \frac{\sigma}{2})-1)^2 + 4\sigma(1+\sigma)(1 - \frac{\hat{\gamma}}{2})\rbrack}^{\frac{1}{2}} \rbrace,
	\end{equation}
	\begin{equation}\label{temperature}
	T = m c^2 \Gamma \frac{v_{sh}}{c}(1 - \frac{\sigma}{2}\frac{c - v_{sh}}{v_sh}),
	\end{equation}
	Where $\sigma$ is the magnetization and $\Gamma$ is the upstream Lorentz-factor measured in the downstream frame.

	The application of the single fluid formalism requires a correct choice of the adiabatic index $\hat{\gamma}$. For a monoatomic gas it is well-defined: in the classical limit $\hat{\gamma} = \frac{5}{3}$ and in the ultra-relativistic limit $\hat{\gamma} = \frac{4}{3}$. However, in the range between these limits the choice of the value of $\hat{\gamma}$ is not trivial. Moreover, the one-fluid model doesn't take into account for the difference between the temperatures of electrons and protons, while the Coulomb relaxation is slow in relativistic astrophysical plasmas. Two-fluid models require some phenomenological approach to account for the collisionless relaxation of the electron temperature.  Therefore, in this study we used the particle-in-cell approach for modelling of relativistic shock waves in order to address the accuracy of MHD equations and to determine the adiabatic index for various cases.

	\section{Numerical setup}
	In this study we performed simulations of a perpendicular relativistic shock  varying the magnetization, Lorentz-factor and the fraction of the turbulent field. 
	The simulation domain is two-dimensional with three-dimensional velocities and fields. To perform such computation we employed the publicly available Tristan-mp code with the explicit numerical scheme developed by Buneman \cite{Buneman93} and  Spitkovsky\cite{Spitkovsky2005}.

	A shock wave is initialized in an electron-proton plasma which is flowing into the simulation box through its right boundary and then is reflecting from the super-conducting wall at the left boundary.  The typical simulation parameters are : the initial flow Lorentz factor $\Gamma = 1.5$, magnetization $\sigma = \frac{B^2}{4\pi\Gamma (n_p m_p + n_e m_e) c^2} = 0.004$ (in the turbulent case $B^2$ is the mean square field). The dimensionless thermal energy $\Delta \gamma = \frac{k T}{m_p c^2}$ is equal to $10^{-4}$ and the electron mass is increased up to $m_e = \frac{m_p}{100}$. The size of the simulation box along the $x$ axis is $L_x = 30000\frac{c}{\omega_p}$ and in the transverse direction $L_y = 400\frac{c}{\omega_p}$, where $\omega_p = \sqrt{\frac{4\pi q^2 n}{\Gamma m_e}}$ is the plasma frequency. These scales correspond to $150000$ and $2000$ grid points in $x$ and $y$ directions, respectively. Also, $2000$ grid points correspond to approximately $10$ gyroradii of the upstream protons.
	In simulations with turbulent fields, we initialized it as  summ of harmonic modes with Kolmogorov's spectum, as described in detail in\cite{Romansky2019}. The ratio of the energy density of the turbulent field to the total magnetic energy density is denoted as $\eta$ , which varied from 0 percent to 90 percent in our simulations. Regular magnetic field is normal to the $x$ axis and is rotated by the angle $\phi$ about $x$ axis: the value $\phi = 0$ corresponds to the out-of-plane orientation of magnetic field and $\phi = 90$ corresponds to the in-plane orientation.

	\section{Results}
	We used several parameter sets, listed in table \ref{setups}, with different magnetization, Lorentz-factors and the fraction of turbulent field. Also, we investigated differences between 2d and 3d cases.
	\begin{table}[h!]
		\label{setups}
		\begin{center}
			\begin{tabular}{|c | c| c | c| c| c| c| c| c| c| c| c| c|}
				\hline
				Setup & Nx & Ny & Nz & $\phi$ & $\sigma$ & $\Gamma$ & $\eta$ & $\frac{v_{sh}}{c}$ & $\hat{\gamma}$ & $T \ 10^{12} K$ & $T_p\ 10^{12}K$ & $T_e \ 10^{12}K$\\
				\hline
				reg1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 0 & 0.355 & 1.513 & 3.53 & 2.5 & 0.23\\
				reg2 & 150000 & 2000 & 1 & 90 & 0.004 & 1.5 & 0 & 0.24 & 1.53 & 2.87 & 2.2 & 0.28\\
				reg3 & 150000 & 2000 & 1 & 0 & 0.0004 & 1.5 & 0 & 0.339 & 1.515 & 3.43 &2.4 & 0.38\\
				reg4 & 150000 & 2000 & 1 & 90 & 0.0004 & 1.5 & 0 & 0.228 & 1.532 & 2.81 &1.65 & 0.37\\
				turb1 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 90 & 0.309 & 1.52 & 3.25 & 2.1 & 0.41\\
				turb2 & 150000 & 2000 & 1 & 0 & 0.004 & 1.5 & 30 & 0.348 & 1.514 & 3.48 & 2.4 & 0.35\\
				gam1 & 150000 & 2000 & 1 & 0 & 0.004 & 2 & 0 & 0.412 & 1.46 & 6.3 & 4.19 & 0.62\\
				gam2 & 150000 & 2000 & 1 & 0 & 0.004 & 5 & 0 & 0.479 & 1.39 & 20.5 & 7.3 & 11.0\\
				gam3 & 150000 & 2000 & 1 & 0 & 0.004 & 10 & 0 & 0.504 & 1.36 & 42.6 & 16.8 & 21.3\\
				reg3d & 80000 & 40 & 40 & 0 & 0.004 & 1.5 & 0 & 0.313 & 1.519 & 3.27 & 3.0 & 0.21\\

				\hline
			\end{tabular}
		\end{center}
		\caption{Parameters of different setups. }
	\end{table}

	Using the density profiles of the simulated shock waves, we determined magnetohydrodynamic parameters of plasma, solving the jump conditions (\ref{hugoniot}). The density averaged in transversal directions has clear jump as shown in figures \ref{density_noturb} and \ref{density_turb}, which is indicating the position of the shock front. So we can evaluate the shock velocity from results of simulations, and then solve jump conditions for the adiabatic index and the downstream temperature, using the iterative method.

	\begin{figure}[h!]
		\centering
		\begin{minipage}{0.49\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{fig/density_noturb.png} 
			\caption{Density profile of a shock wave in a medium with a regular field, averaged through the transversal dimensions, setup reg1.}
			\label{density_noturb}
		\end{minipage}\hfill
		\begin{minipage}{0.5\textwidth}
			\centering
			\includegraphics[width=0.98\textwidth]{fig/density_turb.png} 
			\caption{Density profile of a shock wave in a turbulent medium, averaged through the transversal dimensions, setup turb1.}
			\label{density_turb}
		\end{minipage}
	\end{figure}

	Evaluation of shock front coordinates, which we have determined as a point where  the density is two times larger than the density observed far upstream, at different time steps shows that the shock front velocity is constant in time up to a high accuracy, see figure \ref{shock_x}. Also, it shows, that the shock front velocity decreases with the decreasing magnetization, and, consequently, the adiabatic index does not remain constant. The shock slows down with the increasing turbulent field energy and it requires significant decrease of the adiabatic index. It can be explained by the appearence of new degrees of freedom related to magnetic field oscillations. Obviously, the shock front is faster for higher values of the upstream Lorentz-factor, and the adiabatic index  converges to the value $4/3$ which is the ultra-relativistic limit. For Lorentz-factors less then 10, $\hat{\gamma}$ is significantly higher than the limiting value.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{fig/shock_x.png} 
		\caption{Evolution of the shock front position for various parameter sets.}
		\label{shock_x}
	\end{figure}

	Distributions of particles significantly vary through the x coordinate, and in figure \ref{points} we show the distribution of electrons at different distances from shock front, measured in proton gyroradius $r_g$. Positive distances correspond to upstream region and negative to downstream. As one can see, for distances more than $6 rg$ in the downstream distribution does not depend on coordinate.  We obtained the temperatures of protons and electrons by fitting their distibutions in far downstream with Maxwell-Juttner function, via minimization of the functional $f(T) = \int_{0}^{p_{max}} (F(p) - F_{m}(p,T))^2dp$ by temperature, where $F(p)$ is the electron or proton distribution derived from the simulation, $F_{m}(p,T)$ is the Maxwell-Juttner function with temperature $T$ and $p_{max}$ is the momentum for which the distributioin becomes non-thermal. Obtained temperature has rather week dependence on choice of $p_{max}$. Such evaluation of the downstream temperature is possible even in the cases of strong non-thermal tail, as shown in figure \ref{temp_fit}.

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{fig/electrons_at_points.png} 
		\caption{Distributions of electrons at different regions of simulation box for turb1 parameter set}
		\label{points}
	\end{figure}

	\begin{figure}[h!]
		\centering
		\includegraphics[width=0.7\textwidth]{fig/temperature_fit.png} 
		\caption{Fitting the distribution for turb1 parameter set with Maxwell-Juttner distribution}
		\label{temp_fit}
	\end{figure}

	An analysis of the distribution functions shows that the temperatures of protons and electrons are much smaller than predicted by jump conditions (\ref{hugoniot}). It means that magnetohydrodynamical formalism is not sufficient for the description of downstream temperatures of a relativistic shock wave. One can notice that the ratio of temperature of protons to the temperatures of electrons is not constant: it is close to $\sqrt{\frac{m_p}{m_e}}$ at low Lorenz-factors and decreases down to $1$ at higher ones. Turbulence also has a strong influence on the temperatures, as it makes electrons much hotter, while the protons become colder.

	Finally, a comparison of the results obtained using 2d and 3d setups with the same parameters (setups reg1, reg2 and reg3d) reveals that the differences in results depend on orientation of magnetic field in 2d simulations. For out-of plane orientation these differences are noticeable, but not crucial - temperature and shock velocity differs in 2d and 3d setups by less than on 10 percent. This means that 2d simulations are appropriate for modeling of relativistic collisionless shock waves in magnetized plasmas.

	\section{Conclusions}

	Particle-in-cell simulations have been used to derive effective macroscopic parameters of relativistic shocks in space plasmas. The effective adiabatic index and electron temperatures are two important parameters required to supplement large-scale single-fluid magneto-hydrodynamic simulations of processes in astrophysical plasmas. Further modelling is needed to study the dependences of these two parameters  on the initial conditions which, reflect the presence of MHD large-scale turbulence in the shock upstream.

	\ack
	V I Romansky and A M Bykov acknowledge a support from RSF grant 16-12-10225.
	Results of the work were obtained using computational resources of Peter the Great Saint-Petersburg Polytechnic University Supercomputing Center (http://scc.spbstu.ru)

	\section*{References}
	% \bibliographystyle{apalike}
	\bibliographystyle{iopart-num}

	\bibliography{bibliogr}
\end{document}
